```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# STAT 689 Homework 10
## Rohan Tiwari

### 1. We have talked about the spinal bone mineral density data, with the random intercept case. I stated that I think, or perhaps hope, that the id numbers should not have to be 1,2,3… I also think/hope that the idnumbers do not need to be ordered. It is generally the case that data sets are “cleaned” and there is a code book that converts the actual ids to 1,2,3,…,n. This helps de-identify data and helps preserve anonymity.

### 1a. Test this out in the spinal bone mineral density data, by defining a new variable, femSBMD$idnum2 = 2 * femSBMD$idnum

```{r}
library(HRW)
library(gamm4)
setwd("~/tamu/MS-STAT-2022/689-flexible-regression")
femSBMD = read.csv("femSBMD.csv")
femSBMD$idnum2 = 2*femSBMD$idnum
head(femSBMD)
```

### 1b. Then rerun the gamm4::gamm4 given in class to see if you get the same results. I think/hope you will

```{r}
#original 
origfit <- gamm4(spnbmd ~ s(age,k=10,bs="cr") + black + hispanic + white,
              random= ~(1|idnum),data = femSBMD)

#new
newfit <- gamm4(spnbmd ~ s(age,k=10,bs="cr") + black + hispanic + white,
              random= ~(1|idnum2),data = femSBMD)

summary(origfit$gam)
summary(newfit$gam)
anova(origfit$mer, newfit$mer)
```
Both summary results match. So we can say that the particular value of idnum does not matter in gamm's grouping.

### 2. Get the data set pigWeights.csv from Canvas. The variable weight is the response, and the variable num.weeks is the date of the repeated measures

```{r}
pigs = read.csv(file="pigWeights.csv")
head(pigs)
```

### 2a. Display the lattice plot from library(lattice). Use the example from the spinal bone mineral density data to do this. I covered this in class, but in this case there is only 1 population and no ANCOVA.

```{r}
library(lattice)
pigsvis = xyplot(weight ~ num.weeks,
                     group = id.num,as.table = TRUE,
                     data = pigs,
                     strip = strip.custom(par.strip.text
                                          = list(cex = 1.5)),
                     par.settings = list(layout.heights
                                         =list(strip=1.6)),
                     scales = list(cex = 1.25),
                     xlab = list("num weeks",cex = 1.5),
                     ylab = list(expression(paste(
                       "pig weights")),
                       cex = 1.5),
                     panel = function(x,y,subscripts,groups)
                     {  
                       panel.grid() 
                       panel.superpose(x,y,subscripts,groups,
                                       type = "b",pch = 16,lwd = 2)
                     })
plot(pigsvis)
```

### 2b. Looking at the data, do you think a random intercept model holds for these data? Why or why not? You might want to look at Lecture 15 where I described the means and variances of a random intercept model.

Looking at the plots, the difference seems to be the starting point but the slope seems to be the same which indicatesthat there is a random intercept (the starting point for each plot seems random)

### 2c. Fit the random intercept model with num_weeks modeled as a spline. Do a summary and show your results. Show the between-person variance of the intercept and the within-person variance of the random errors. You may use either mgcv::gamm or gamm4::gamm4.

```{r}
fitpigs <- gamm4(weight ~ s(I(num.weeks),k=8,bs="cr"),
             random= ~(1|id.num),data = pigs)

summary(fitpigs$gam)
summary(fitpigs$mer)
```
sigma_u^2 is 15.1524
sigma_e^2 is 4.2981


### 2d. Using anova() in gamm4::gamm4, to test whether a spline is needed as compared to a linear and a quadratic effect.

```{r}
fitL <- gamm4(weight ~ I(num.weeks),
              random= ~(1|id.num),data = pigs)
summary(fitL$gam)
fitQ <- gamm4(weight ~ I(num.weeks) + I(num.weeks^2),
              random= ~(1|id.num),data = pigs)
summary(fitQ$gam)
#compare fits
anova(fitpigs$mer,fitL$mer)
anova(fitpigs$mer,fitQ$mer)
```
Linear versus spline: The p-value is 0.09679 which indicates that spline model is not significant. Quadratic versus Spline: Does not produce a p-value.

### 2e. Compare the quadratic and linear fits as well

```{r}
anova(fitQ$mer,fitL$mer)
```
The p-value is 0.3869 which means quadratic model is not significant. So overall, we can see that linear model might be the best fit here. 


### 2f. As in the spinal bone mineral density data, plot the fixed effects function against num.weeks, and include a pointwise 95% confidence interval for it.

```{r}
plot(fitpigs$gam,shade = TRUE,shade.col = "palegreen",bty = "l")
```



### 2g. Since you have already computed var(U) and var( ε ), what is the estimated withinperson correlation for this model?

```{r}
#sigma_u^2 is 15.1524
#sigma_e^2 is 4.2981
ratio = 4.2981/15.1524
corr = (1 + ratio) ^ -1
print("Within person correlation is: ")
corr
```

